---
title: "Wind value rasterization"
output: Raster containing wind-related values
date: "2025-12-01"
---
This script rasterizes wind-related variables for the Canary Islands, including mean wind speed (`SPD40`) and wind turbulence (`WEIBK`), converting point shapefile data into raster grids at a 100 m resolution. The resulting rasters can be used for spatial analysis or as input for species distribution models and environmental modeling.

Load libraries
```{r}
library(terra)
library(sf)

```

Load data
```{r load shp file}
island <- "GranCanaria" # Other islands: LaPalma, LaGomera, Tenerife, ElHierro
shp <- st_read(paste0("pathtofile/SHP40_", island, ".shp"))
```

Rasterize mean wind speed (m/s)
```{r}
r <- rast(shp, resolution = 100)  #create empty raster
wind_speed <- rasterize(shp, r, field = "SPD40")
writeRaster(wind_speed, paste0(island, "_speed.tif"), overwrite = TRUE)
```

Rasterize wind turbulence (Weibk distribution)
```{r}
r <- rast(shp, resolution = 100)  #create empty raster
weibk <- rasterize(shp, r, field = "WEIBK")
writeRaster(weibk, paste0(island, "_weibk.tif"), overwrite = TRUE)
```



