---
title: "Create raster input"
description: short R code to create raster inputs for ResistanceGA
output: Raster for resistance surface modeling
---


Load libraries
```{r setup, include=FALSE, message=FALSE}
# Cargar paquetes esenciales para modelado, visualización y manejo de datos espaciales
library(biomod2)     # Modelado de distribución de especies
library(terra)       # Manejo de datos raster
library(sp)          # Datos espaciales
library(raster)      # Manipulación de datos raster
library(sf)          # Datos espaciales vectoriales

# Paquetes para análisis y manipulación de datos
library(dplyr)
library(tidyr)
library(tidyverse)
library(gtools)
library(Hmisc)
library(caret)

# Paquetes de visualización
library(ggplot2)
library(gridExtra)
library(viridis)
library(RColorBrewer)
library(rasterVis)
library(adegraphics)

# Análisis estadístico y modelado
library(ade4)
library(adehabitatHS)
library(gam)
library(earth)
library(rpart)
library(mda)
library(nnet)
library(randomForest)
library(gbm)
library(MASS)
library(pROC)

# Otros paquetes
library(here)
library(knitr)
library(rgbif)
library(corrgram)
library(reshape2)
library(readxl) #read excel files


```




Load climate + topography data
The raster is a raster stack that contains the input layers that you would like to consider for analysis in ResistanceGA. Otherwise, load rasters individually and merge with terra package. 
```{r }
raster <- rast("Spatraster.tif")

```


Optional: Crop to island size

Our original rasters are Canary Islands, but we do island-specific analyses individually. therefore, this code will crop the rasters to the vector that you chose, in our case we crop it to a vector showcasing the islands limits. 
Here we showcase the example of bio1, but we did this for other rasters that we took into consideration such as bio2, bio12, bio18, slope, profile curvature, wind rasters.
```{r}
vect <- vect(paste0("islandcontorno.shp"))
b1 <- crop(raster$bio1, vect)

```

Save raster
```{r}
writeRaster(b1, ".raster1.asc", filetype = "AAIGrid", overwrite=TRUE)

```


